---
title: "Schedule"
description: 
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

Below is our course schedule, which may change as the course progresses. 

The morning session will run from 10:00 am-12:00 pm and the afternoon session will run from 1:30-3:30 pm (all times EDT/Toronto time).  

Some days will include a lunchtime guest speaker from DoSS or the university from 12:15-1:15 pm.  A list of guest speakers is given below.

All sessions will use the following zoom meeting:


## Class schedule

```{r, echo=FALSE}
library(tidyverse)
library(kableExtra)
options(knitr.kable.NA = '')
my_backgrounds <- colorRampPalette(c("white", "#002A5C"))(5)
k <- 9
row_nums <-sort(c(seq(1,31,4), seq(2,31,4)))
jsc270courseschedule <- read_csv("Schedule/schedule.csv")
knitr::kable(jsc270courseschedule[1:31,]) %>%  
  kable_styling(full_width = T, position = "float_left")   %>%   
  row_spec(row_nums ,background = my_backgrounds[2]) 
```

## Guest speaker schedule

```{r, echo=FALSE}
my_backgrounds <- colorRampPalette(c("white", "#002A5C"))(5)
row_nums <-c(1,2,3,4,10,11)
jsc270courseschedule <- read_csv("Schedule/guest_speaker_schedule.csv")
knitr::kable(jsc270courseschedule[1:12,]) %>%  
  kable_styling(full_width = T, position = "float_left")   %>%   
  row_spec(row_nums ,background = my_backgrounds[2]) 
```
