---
title: 'Solution 4: Statistical inference (I)'
author: "Jianhui Gao"
date: \today
output: pdf_document
---

# Part 1: Probability distributions

Question a: A contestant on a game show needs to answer 10 questions correctly to win the jackpot. However, if they get 4 incorrect answers, they are kicked off the show. Suppose one contestant consistently has a 80\% chance of correctly responding to any question. 

\begin{enumerate}
\item What is the probability distribution?
\item What is the probability that she will correctly answer 10 questions before 4 incorrect responses? 
\item Write out the R code to calculate (b).
\end{enumerate}

**Solution** 

1.  The experiment is modeled by the \textbf{Negative Binomial Distribution}, as it counts the number of failures ($k$) before a specified number of successes ($r$) occurs.
    \begin{itemize}
        \item Success: Correct answer, $p = 0.8$
        \item Failure: Incorrect answer, $q = 1-p = 0.2$
        \item Required successes: $r=10$
        \item Losing condition: $k=4$ failures
    \end{itemize}
2. The contestant wins if she achieves $r=10$ successes before getting 4 failures. This means the total number of failures $k$ must be in the set $\{0, 1, 2, 3\}$. The probability mass function is $P(X=k) = \binom{k+r-1}{r-1} p^r q^k$. The total probability of winning is the sum over the allowed values of $k$:
    $$ P(\text{Win}) = \sum_{k=0}^{3} \binom{k+10-1}{10-1} (0.8)^{10} (0.2)^k $$
$$ P(\text{Win}) = \binom{9}{9}(0.8)^{10}(0.2)^0 + \binom{10}{9}(0.8)^{10}(0.2)^1 + \binom{11}{9}(0.8)^{10}(0.2)^2 + \binom{12}{9}(0.8)^{10}(0.2)^3 $$
    $$ P(\text{Win}) \approx 0.1074 + 0.2147 + 0.2362 + 0.1890 \approx 0.7473 $$

3. The calculation can be performed concisely using R's cumulative negative binomial function, \texttt{pnbinom}.

\begin{verbatim}
# Calculates the probability of 0, 1, 2, or 3 failures (q)
# before 10 successes (size) are achieved.
pnbinom(q = 3, size = 10, prob = 0.8)

# Output: [1] 0.747323
\end{verbatim}

Question b: A small town’s police department issues 5 speeding tickets per month on average. 

\begin{enumerate}
\item Using a Poisson random variable, what is the likelihood that the police department issues 3 or fewer tickets in one month? 
\item What is the probability that 10 days or fewer elapse between two tickets being issued?
\item Write out the R code to calculate (a), (b).
\end{enumerate}

**Solution** 

1. First, we note that here $P(Y \leq 3)=P(Y=0)+P(Y=1)+\cdots+P(Y=3)$.
Applying the probability mass function for a Poisson distribution with $\lambda=5$, we find that

$$
\begin{aligned}
P(Y \leq 3) &=P(Y=0)+P(Y=1)+P(Y=2)+P(Y=3) \\
&=\frac{e^{-5} 5^{0}}{0 !}+\frac{e^{-5} 5^{1}}{1 !}+\frac{e^{-5} 5^{2}}{2 !}+\frac{e^{-5} 5^{3}}{3 !} \\
&=0.27 .
\end{aligned}
$$

```{r}
# q is the number of events (3 or fewer).
# lambda is the average rate of events per interval.
ppois(q = 3, lambda = 5)
```

2. We know the town's police issue 5 tickets per month. For simplicity's sake, assume each month has 30 days. Then, the town issues $\frac{1}{6}$ tickets per day. That is $\lambda=\frac{1}{6}$, and the average wait time between tickets is $\frac{1}{1 / 6}=6$ days. Therefore,
$$
P(Y<10)=\int_{0}^{10} \frac{1}{6} e^{-\frac{1}{6} y} d y=0.81
$$

```{r}
pexp(10, rate = 1/6)
```

# Part 2: Statistical inference

\begin{enumerate}
\item (AoS 6.6.2) Let $X_{1}, \ldots, X_{n} \sim \operatorname{Uniform}(0, \theta)$ and let $\hat{\theta}=\max \left\{X_{1}, \ldots, X_{n}\right\}$. Find the bias, se and MSE of this estimator.

\item (AoS 6.6.3) Let $X_{1}, \ldots, X_{n} \sim \operatorname{Uniform}(0, \theta)$ and let $\hat{\theta}=2 \bar{X}_{n}$. Find the bias, se and MSE of this estimator.

\item Let $X_{1}, \ldots, X_{n} \sim \operatorname{Uniform}(0,1)$. Let $Y_{n}=\bar{X}_{n}^{2}$. Find the limiting distribution of $Y_{n}$. (Hint: CLT)

\end{enumerate}

**Solution** 

1. The CDF $G$ of $\widehat{\theta}$ is
$$
\begin{aligned}
G(\widehat{\theta}) &=\mathbb{P}(\widehat{\Theta} \leq \widehat{\theta}) \\
&=\mathbb{P}\left(\max \left\{X_{1}, \ldots, X_{n}\right\} \leq \widehat{\theta}\right) \\
&=\prod_{i=1}^{n} \mathbb{P}\left(X_{i} \leq \widehat{\theta}\right) \\
&=F_{\theta}(\widehat{\theta})^{n} \\
&=\left(\frac{\widehat{\theta}}{\theta}\right)^{n}
\end{aligned}
$$
The density is therefore
$$
g(\widehat{\theta})=\left(\frac{n}{\theta}\right)\left(\frac{\widehat{\theta}}{\theta}\right)^{n-1}
$$
Thus,
$$
\mathbb{E}_{\theta}(\widehat{\theta})=\int_{0}^{\theta} \widehat{\theta} g(\widehat{\theta}) d \widehat{\theta}=\frac{n \theta}{n+1}
$$
and
$$
\text { bias }=\frac{n \theta}{n+1}-\theta=-\frac{\theta}{n+1} .
$$
Also,
$$
\mathbb{E}_{\theta}\left(\widehat{\theta}^{2}\right)=\int_{0}^{\theta} \widehat{\theta}^{2} g(\widehat{\theta}) d \widehat{\theta}=\frac{n \theta^{2}}{n+2}
$$
and so
$$
\mathbb{V}_{\theta}(\widehat{\theta})=\frac{n \theta^{2}}{n+2}-\left(\frac{n \theta}{n+1}\right)^{2}=\frac{n \theta^{2}}{(n+2)(n+1)^{2}}
$$
The mse is
$$
\operatorname{bias}^{2}+\mathbb{V}=\left(\frac{\theta}{n+1}\right)^{2}+\frac{n \theta^{2}}{(n+2)(n+1)^{2}}=\frac{2 \theta^{2}}{(n+1)(n+2)}
$$

2. Recall that $\mathbb{E}\left(X_{i}\right)=\theta / 2, \mathbb{V}\left(X_{i}\right)=\theta^{2} / 12$. So
$$
\mathbb{E}_{\theta}(2 \bar{X})=2 \mathbb{E}_{\theta}(\bar{X})=2 \frac{\theta}{2}=\theta
$$
and hence bias $=0$. Now
$$
\mathbb{V}_{\theta}(2 \bar{X})=4 \mathbb{V}_{\theta}(\bar{X})=\frac{4 \sigma^{2}}{n}=\frac{4 \theta^{2}}{12 n}=\frac{\theta^{2}}{3 n} .
$$
Since this estimator is unbiased,
$$
\mathrm{mse}=\mathbb{V}_{\theta}(\widehat{\theta})=\frac{\theta^{2}}{3 n} .
$$

3. $\mu=\mathbb{E}\left(X_{i}\right)=1 / 2$ and $\sigma^{2}=\mathbb{V}\left(X_{i}\right)=1 / 12$. By the CLT,
$$
\frac{\sqrt{n}(\bar{X}-\mu)}{\sigma}=\sqrt{12 n}\left(\bar{X}-\frac{1}{2}\right) \rightsquigarrow N(0,1) .
$$
Now $Y=g(\bar{X})$ where $g(s)=s^{2}$. And $g^{\prime}(s)=2 s$ and $g^{\prime}(\mu)=g^{\prime}(1 / 2)=$ $2(1 / 2)=1$. From the delta method,
$$
\frac{\sqrt{n}(Y-g(\mu))}{\left|g^{\prime}(\mu)\right| \sigma}=\sqrt{12 n}\left(\bar{X}-\frac{1}{4}\right) \leadsto N(0,1)
$$
