\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{longtable}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
}

%\title{DoSS Summer Prep Bootcamp: Operational math}
%\author{jgronsbell }
\date{ }

\begin{document}

\section*{Operational Math\\  {\it{DoSS Summer Prep Bootcamp 2022}}}

\section{Time \& Place}
TBD

\section{Instructor}
\href{https://www.statistics.utoronto.ca/people/directories/graduate-students/emma-kroell}{Emma Kroell}, 2nd year PhD student, Department of Statistical Sciences


\section{Course Outline}
Review of proof techniques. Selected topics in linear algebra, real analysis and topology.

\section{Textbooks}
The following books are optional texts for the different areas we will cover. All books are freely available online, however some may require a U of T log-in.

% The Tools of Mathematical Reasoning

\noindent
{\bf{Linear algebra}}:

\vspace{0.1cm}

1. \href{https://link-springer-com.myaccess.library.utoronto.ca/book/10.1007/978-3-319-11080-6}{{\emph{Linear Algebra Done Right}}} by Sheldon Axler

2. \href{https://www.math.brown.edu/streil/papers/LADW/LADW.html}{{\emph{Linear Algebra Done Wrong}}} by Sergei Treil

\noindent
{\bf{Analysis}}:
\vspace{0.1cm}

3. \href{https://digitalcommons.trinity.edu/mono/7/}{{\emph{Introduction to Real Analysis}}} by William F. Trench

4. \href{https://link-springer-com.myaccess.library.utoronto.ca/book/10.1007/978-1-4939-2712-8}{{\emph{Understanding Analysis}}} by Stephen Abbott

5. \href{https://link-springer-com.myaccess.library.utoronto.ca/book/10.1007/978-3-319-17771-7}{\emph{Real Mathematical Analysis}} by Charles C. Pugh

6. \href{https://link-springer-com.myaccess.library.utoronto.ca/book/10.1007/0-387-28387-0}{\emph{A Taste of Topology}} by Volker Runde


\vspace{0.4cm}

\noindent
{\bf{Additional resources}}:
\vspace{0.1cm}

\href{http://84.89.132.1/~piotr/docs/RealAnalysisNotes.pdf}{\emph{Lecture notes in Mathematics for Economics and Statistics}} by Piotr Zwiernik

\href{http://www.math.uwaterloo.ca/~krdavids/PM351/PMath351Notes.pdf}{{\emph{Real Analysis Lecture Notes}}} by Ken Davidson

\href{http://www.math.uwaterloo.ca/~lwmarcou/notes/pmath351.pdf}{{\emph{Real Analysis Lecture Notes}}} by Laurent Marcoux

\href{http://www.math.uwaterloo.ca/~krdavids/FA/PM453Notes.pdf}{{\emph{Functional Analysis Lecture Notes}}} by Ken Davidson




% add some comment on which chapter



\newpage
\section{Tentative Lecture Schedule}
The lecture topics and corresponding chapters in the texts (if applicable) are outlined below.  This schedule is tentative and will inevitably be augmented during the course.  


\begin{longtable}{| p{1.5cm}  | p{8cm} | p{3cm} |}
\hline
{\bf{Lecture}}  & {\bf{Topics}} &  {\bf{References}} \\
\hline
1  & Review of logic \& proof techniques &   \\
\hline
2 &  Linear algebra I & 1 \& 2 \\
\hline
3  &  Linear algebra II & 1 \& 2  \\
\hline
4  & Linear algebra III &  1 \& 2 \\
\hline
5  & Set theory & 6 (ch. 1) \\
\hline
6  &  Metric spaces and sequences I &  3, 4, 5\\
\hline
7  & Metric spaces and sequences II & 3, 4, 5, 6 (ch. 2)  \\
\hline
8  & Topology & 6 (ch. 3 \& 4) \\
 \hline
9  &  Differentiation and integration &  3, 4, 5 \\
\hline
10  &   Multivariable calculus & 3, 4, 5\\
\hline
\end{longtable}







\end{document}

