---
title: 'Solution 1: Basic programming in R'
author: "Yaqi Shi"
date: "07/09/2024"
output: pdf_document
---

# Part 1: Matrix and vector operations. 

1. Solve the following system: 

$$
\left[\begin{array}{ccccc}
a_1 & b_1 &  & & 0 \\
c_1 & a_2 & b_2 & & \\
& \ddots & \ddots & \ddots & \\
& &  & a_{99} & b_{99} \\
0 & & & c_{99} & a_{100} 
\end{array}\right] 
\left[\begin{array}{c}
x_{1} \\
x_{2} \\
\vdots \\
x_{100}
\end{array}\right]
=
\left[\begin{array}{c}
d_{1} \\
d_{2} \\
\vdots \\
d_{100}
\end{array}\right]
$$
where 
$$
a_j = j, \quad b_j = 1/j, \quad c_j = 1, \quad d_j = \sin(j \pi / 200)
$$
and print $x_1, x_2, \dots, x_5$. 

**Solution**:
```{r}
# Define A.
A <- matrix(rep(0, 100*100), nrow = 100, ncol = 100, byrow = TRUE)

for (i in c(1:100)) {
    A[i, i] <- i

    if (i + 1 < 101)  A[i, i + 1] <- 1/i
    if (i - 1 > 0)    A[i, i - 1] <- 1
}

# Define D.
d <- c(1:100)
d <- sin(d*pi/200)

# Solve Ax = d.
x <- solve(A, d)
x[1:5]
```

# Part 2: For loops. 

1. Write a function that uses a `for` loop to calculate the following with a sequence of $m$, and generate a plot for $m$ verses $E_m$. Avoid using a `for` loop, can you complete the same task?  

$$
E_m = 1 + \frac{1}{2} + \dots + \frac{1}{2^m} - \log(2^m)
$$

**Solution**:
```{r}
# Using for loop.
E_m <- function(m) {
  res <- 1
  for (i in c(1:m)) {
    res <- res + 1/(2^i)
  }
  res <- res - log(2^m)
  
  return(res)
}

# Avoid using for loop.
E_m2 <- function(m) {
  res <- 1
  
  index <- c(1:m)
  denom <- 2^index
  res <- res + sum(1/denom)

  res <- res - log(2^m)
  
  return(res)
}

E_m(100)
E_m2(100)
```

\clearpage

2. Let's draw a regular polygon of $n$ sides, with a horizontal bottom side, and the corners of the polygon staying on the unit circle. For given $n = 5$ and $r = 0.6$, start the first point $p_1 = (x_1, y_1)$ as a random number in $(-0.5, 0.5) \times (-0.5, 0.5)$, generate $10^4$ points interactively. In the $j$th iteration, we choose one corner $z_*$ of the polygon randomly and let $p_{j+1} = (x_{j+1}, y_{j+1})$ be the point on the line segment between $p_j$ and $z_*$, with the distance from $p_{j+1}$ to $p_j$ being $r$ times the distance from $z_j$ to $p_j$ and then draw all these points as dots in the $xy$ panel. 

Hint: Complex numbers can be used to represent points in the $xy$ plane. The following script works for even or odd $n$, and the polygon always has a flat bottom. 

```{r, fig.width=5}
n <- 5
t <- c(0:n) - 0.5
z <- exp(2i * pi * (t/n - 0.25))


plot(Re(z), Im(z))
```

**Solution**:
```{r, fig.width=5}
n_points <- 10000

# generate 10000 points
p <- rep(NA, n_points)

# starting from the 1st point from [-0.5, 0.5]x[-0.5, 0.5]
rand <- runif(1, min = 0, max = 1)
p[1] <- rand - 0.5 + 1i * (rand - 0.5)

# r x distance of d(z_j, p_j)
r <- 0.6

# set up the n sides polygon and n corners
n <- 5
t <- c(0:n) - 0.5
z <- exp(2i * pi * (t/n - 0.25))

# in the jth iteration, randomly choose one corner and update p(j+1)
for (j in c(2:n_points)) {
  # sample 1 corner
  zs <- z[sample(n, 1)]
  p[j] <- p[j-1] + r*(zs - p[j-1])
}
  
plot(Re(p), Im(p))
```






