---
title: 'Solution 4: Statistical inference (I)'
author: "Siyue Yang"
date: "05/10/2022"
output: pdf_document
---

# Part 1: Probability distributions

\begin{enumerate}
\item A contestant on a game show needs to answer 10 questions correctly to win the jackpot. However, if they get 4 incorrect answers, they are kicked off the show. Suppose one contestant consistently has a 80\% chance of correctly responding to any question. 

\begin{enumerate}
\item What is the probability distribution?
\item What is the probability that she will correctly answer 10 questions before 4 incorrect responses? 
\item Write out the R code to calculate (b).
\end{enumerate}

\item A small town’s police department issues 5 speeding tickets per month on average. 

\begin{enumerate}
\item Using a Poisson random variable, what is the likelihood that the police department issues 3 or fewer tickets in one month? 
\item What is the probability that 10 days or fewer elapse between two tickets being issued?
\item Write out the R code to calculate (a), (b).
\end{enumerate}

\end{enumerate}

**Solution** 

1. Negative Binomial distribution.

Letting $Y$ represent the number of incorrect responses, and setting $r=10$, we want
$$
\begin{aligned}
P(Y<4)=& P(Y=0)+P(Y=1)+P(Y=2)+P(Y=3) \\
=&\left(\begin{array}{c}
9 \\
9
\end{array}\right)(1-0.8)^{0}(0.8)^{10}+\left(\begin{array}{c}
10 \\
9
\end{array}\right)(1-0.8)^{1}(0.8)^{10} \\
&+\left(\begin{array}{c}
11 \\
9
\end{array}\right)(1-0.8)^{2}(0.8)^{10} + \left(\begin{array}{c}
12 \\
9
\end{array}\right)(1-0.8)^{3}(0.8)^{10}\\
=& 0.97
\end{aligned}
$$

```{r}
sum(dnbinom(0:3, size = 10, prob = .8))
```

2. First, we note that here $P(Y \leq 3)=P(Y=0)+P(Y=1)+\cdots+P(Y=3)$.
Applying the probability mass function for a Poisson distribution with $\lambda=5$, we find that
$$
\begin{aligned}
P(Y \leq 3) &=P(Y=0)+P(Y=1)+P(Y=2)+P(Y=3) \\
&=\frac{e^{-5} 5^{0}}{0 !}+\frac{e^{-5} 5^{1}}{1 !}+\frac{e^{-5} 5^{2}}{2 !}+\frac{e^{-5} 5^{3}}{3 !} \\
&=0.27 .
\end{aligned}
$$

```{r}
sum(dpois(0:3, lambda = 5))   # or use ppois(3, 5)
```

We know the town's police issue 5 tickets per month. For simplicity's sake, assume each month has 30 days. Then, the town issues $\frac{1}{6}$ tickets per day. That is $\lambda=\frac{1}{6}$, and the average wait time between tickets is $\frac{1}{1 / 6}=6$ days. Therefore,
$$
P(Y<10)=\int_{0}^{10} \frac{1}{6} e^{-\frac{1}{6} y} d y=0.81
$$

```{r}
pexp(10, rate = 1/6)
```

# Part 2: Statistical inference

\begin{enumerate}
\item (AoS 6.6.2) Let $X_{1}, \ldots, X_{n} \sim \operatorname{Uniform}(0, \theta)$ and let $\hat{\theta}=\max \left\{X_{1}, \ldots, X_{n}\right\}$. Find the bias, se and MSE of this estimator.

\item (AoS 6.6.3) Let $X_{1}, \ldots, X_{n} \sim \operatorname{Uniform}(0, \theta)$ and let $\hat{\theta}=2 \bar{X}_{n}$. Find the bias, se and MSE of this estimator.

\item Let $X_{1}, \ldots, X_{n} \sim \operatorname{Uniform}(0,1)$. Let $Y_{n}=\bar{X}_{n}^{2}$. Find the limiting distribution of $Y_{n}$. (Hint: CLT)

\end{enumerate}

**Solution** 

1. The CDF $G$ of $\widehat{\theta}$ is
$$
\begin{aligned}
G(\widehat{\theta}) &=\mathbb{P}(\widehat{\Theta} \leq \widehat{\theta}) \\
&=\mathbb{P}\left(\max \left\{X_{1}, \ldots, X_{n}\right\} \leq \widehat{\theta}\right) \\
&=\prod_{i=1}^{n} \mathbb{P}\left(X_{i} \leq \widehat{\theta}\right) \\
&=F_{\theta}(\widehat{\theta})^{n} \\
&=\left(\frac{\widehat{\theta}}{\theta}\right)^{n}
\end{aligned}
$$
The density is therefore
$$
g(\widehat{\theta})=\left(\frac{n}{\theta}\right)\left(\frac{\widehat{\theta}}{\theta}\right)^{n-1}
$$
Thus,
$$
\mathbb{E}_{\theta}(\widehat{\theta})=\int_{0}^{\theta} \widehat{\theta} g(\widehat{\theta}) d \widehat{\theta}=\frac{n \theta}{n+1}
$$
and
$$
\text { bias }=\frac{n \theta}{n+1}-\theta=-\frac{\theta}{n+1} .
$$
Also,
$$
\mathbb{E}_{\theta}\left(\widehat{\theta}^{2}\right)=\int_{0}^{\theta} \widehat{\theta}^{2} g(\widehat{\theta}) d \widehat{\theta}=\frac{n \theta^{2}}{n+2}
$$
and so
$$
\mathbb{V}_{\theta}(\widehat{\theta})=\frac{n \theta^{2}}{n+2}-\left(\frac{n \theta}{n+1}\right)^{2}=\frac{n \theta^{2}}{(n+2)(n+1)^{2}}
$$
The mse is
$$
\operatorname{bias}^{2}+\mathbb{V}=\left(\frac{\theta}{n+1}\right)^{2}+\frac{n \theta^{2}}{(n+2)(n+1)^{2}}=\frac{2 \theta^{2}}{(n+1)(n+2)}
$$

2. Recall that $\mathbb{E}\left(X_{i}\right)=\theta / 2, \mathbb{V}\left(X_{i}\right)=\theta^{2} / 12$. So
$$
\mathbb{E}_{\theta}(2 \bar{X})=2 \mathbb{E}_{\theta}(\bar{X})=2 \frac{\theta}{2}=\theta
$$
and hence bias $=0$. Now
$$
\mathbb{V}_{\theta}(2 \bar{X})=4 \mathbb{V}_{\theta}(\bar{X})=\frac{4 \sigma^{2}}{n}=\frac{4 \theta^{2}}{12 n}=\frac{\theta^{2}}{3 n} .
$$
Since this estimator is unbiased,
$$
\mathrm{mse}=\mathbb{V}_{\theta}(\widehat{\theta})=\frac{\theta^{2}}{3 n} .
$$

3. $\mu=\mathbb{E}\left(X_{i}\right)=1 / 2$ and $\sigma^{2}=\mathbb{V}\left(X_{i}\right)=1 / 12$. By the CLT,
$$
\frac{\sqrt{n}(\bar{X}-\mu)}{\sigma}=\sqrt{12 n}\left(\bar{X}-\frac{1}{2}\right) \rightsquigarrow N(0,1) .
$$
Now $Y=g(\bar{X})$ where $g(s)=s^{2}$. And $g^{\prime}(s)=2 s$ and $g^{\prime}(\mu)=g^{\prime}(1 / 2)=$ $2(1 / 2)=1$. From the delta method,
$$
\frac{\sqrt{n}(Y-g(\mu))}{\left|g^{\prime}(\mu)\right| \sigma}=\sqrt{12 n}\left(\bar{X}-\frac{1}{4}\right) \leadsto N(0,1)
$$
